---
title: "sta206hw7"
author: "RanLiu"
date: "November 18, 2015"
output: word_document
---

# 2.Processing of the data

## 2a.
```{r}
setwd("E:/RLiu_R/sta206")
db = read.table("diabetes.txt", header = TRUE)
options(width = 60)
head(db)
table(db$frame)
db$frame[db$frame == ""] = NA
db$frame = droplevels(db$frame)
table(db$frame)
```

## 2b.
```{r}
names(db)
drops = c("id", "bp.2s", "bp.2d")
db = db[, !names(db) %in% drops]
names(db)
```

## 2c.
```{r}
#The following variables are quantitative variables:
(qt = names(db)[!sapply(db, is.factor)])
#The following variables are qualitative variables:
(ql = names(db)[sapply(db, is.factor)])
#histogram for glyhb
hist(db$glyhb)
```

The distribution of glyhb is severely right-skewed.

Histogram for the other 12 quantitative variables:
```{r}
other.qt = names(db)[names(db) %in% qt & names(db)!= "glyhb"]
par(mfrow = c(3,4), mar = c(2, 2, 2, 2))
invisible(sapply(other.qt, function(x) hist(db[,names(db) == x], main = x)))
```

Pie charts for qualitative variables:
```{r}
par(mfrow = c(1,3), mar = c(2, 2, 2, 2))
invisible(sapply(ql, function(x) pie(table(db[,names(db) == x]),main = x)))
```

## 2d.
```{r}
gl = db$glyhb
tgl = data.frame("log.glyhb" = log(gl), "sqrt.glyhb" = sqrt(gl) , "revert.glyhb" = 1/gl)
par(mfrow = c(2,2))
invisible(sapply(c(1:3), function(x) hist(tgl[,x], xlab = names(tgl)[x], main = paste0("histogram of ", names(tgl)[x]))))
```

None of the three looks normal like to me... the log and squareroot are still right-skewed and the reverted is left-skewed. If I have to choose one, then maybe reverted glyhb is the most Normal-like one, among the three.

## 2e.
```{r}
db$glyhb.star = (1/(db$glyhb))
db = db[, !names(db) == "glyhb"]
```

## 2f.
```{r}
index.na=apply(is.na(db), 1, any)
## identify cases with missing value.
db.s=db[index.na == FALSE,] ##drop cases with missing value.
any(is.na(db.s)) ## this should return FALSE -- no NA in db.s
dim(db.s) ##this should return 366 16: 366 cases, 16 variables.
table(db.s$frame) ## this should show three classes.
```

## 2g.
```{r}
par(mfrow = c(1,1), mar = c(0,0,0,0))
plot(db.s[,names(db.s) %in% qt], pch = ".")
correlation = cor(db.s[,names(db.s) %in% qt])
round(correlation, 2)
```

Nonlinearity is not observed from scattered plot.

## 2h.
```{r}
par(mar = c(2, 2, 2, 2))
boxplot(db$glyhb.star ~ db$gender, col = c("pink", "blue"), main = "glyhb vs gender")
boxplot(db$glyhb.star ~ db$frame, col = c("red", "orange", "yellow"), main = "glyhb vs frame")

```

## 2i.
```{r}
set.seed(206) ## set seed for random number generator
##so everyone gets the same split of the data.
n.s=nrow(db.s) ## number of cases in db.s (366)
index.s=sample(1: n.s, size=n.s/2, replace=FALSE)
## randomly sample 183 cases to form the training data.
db.t=db.s[index.s,] ## get the training data set.
## the remaining 183 cases form the validation set.
db.v=db.s[-index.s,]
```

## 2j.
```{r}
compare = function(varname){
  boxplot(db.t[, varname], db.v[, varname], main = varname, names = c("train", "valid"), col = c("green", "orange"))
}
vars = c("glyhb.star", "stab.glu", "ratio", "age", "bp.1s", "waist")
par(mfrow = c(1,2))
invisible(lapply(vars, compare))
par(mfrow = c(1,1))
```

Yes,  these variables having similar distributions in these two sets.

# 3.Selection of first-order effects

## 3a.
```{r}
model1 = lm(glyhb.star ~ ., data = db)
anova(model1)["Residuals", 3]
library(MASS)
boxcox(glyhb.star ~ ., data = db)
```

* There are 16 regression coefficients in this model. 
* MSE is  0.001326992.
* We can see the log-likelihood is largest when λ is around 1, which means no additional transformation is needed.

## 3b.
```{r}
library(leaps)
sub = regsubsets(x = glyhb.star ~ ., data = db, nbest = 1, nvmax = 16)
sum_sub = summary(sub)
p.m = as.integer(rownames(sum_sub$which))+1
#number of coefficients in each model: p.m
ssto = sum((db.t$glyhb.star-mean(db.t$glyhb.star))^2)
sse = (1-sum_sub$rsq)*ssto
n = nrow(db.t)
aic = n*log(sse/n)+2*p.m
bic = n*log(sse/n)+log(n)*p.m
res_sub = data.frame(p.m, sse, rsq = sum_sub$rsq, adjr2 = sum_sub$adjr2, cp = sum_sub$cp, aic, bic, sum_sub$which)
# index of best models under 'sse', 'aic', 'bic' criteria:
index1 = apply(res_sub[, c('sse', 'aic', 'bic')], 2, which.min)
# index of best models under 'rsq', 'adjr2' criteria:
index2 = apply(res_sub[, c('rsq', 'adjr2')], 2, which.max)
# best models under cp criteria:
par(mar = c(4 ,4, 4, 4))
plot(p.m, res_sub$cp, xlab="p", ylab="Cp", pch=19, main = "Best model under Cp criteria")
abline(0,1,col="red")
index3 = 5
```

The best model under Cp criteria should be one which has nearest cp value to p and a small cp. So p = 5, cp = 5.077535 is the best one.

Now put them together to see which varibles each model includes.
```{r}
# put them together
index = c(index1, index2, index3)
best_models = sum_sub$which[index, ]
rownames(best_models) = c('sse', 'aic', 'bic','rsq', 'adjr2','cp')
(best_models = apply(best_models, 1, function(x) colnames(best_models)[x]))
```
 
## 3c.
```{r}
none_mod = lm(glyhb.star ~ 1, data=db.t)
full_mod1 = lm(glyhb.star ~ ., data=db.t)
modelfs1 = stepAIC(none_mod, scope=list(upper=full_mod1), direction="both", k=2)
modelfs1$anova
best_models$aic
min(res_sub$aic)
```

The best model got in previous question includes stab.glu + ratio + age + waist + time.ppn" with AIC = -1200.249.

The best model under forward step-wise method includes stab.glu + ratio + age + location with AIC = -1176.62.

The best model under forward step-wise method is larger than that of the best model got in previous question. So forward step-wise probabaly gives a suboptimal model.

## 3d.
```{r, "residual vs. fitted value plot"}
plot(lm(formula = glyhb.star ~ stab.glu + ratio + age + location, data = db.t), which = 1)
```

Yes, modelfs1 seems adequate. There seems to be a little bit residual variance inequinity though.

# 4.Selection of first- and second- order effects
## 4a.
```{r}
model2 = lm(glyhb.star ~ (.)^2, data = db.t)
anova(model2)["Residuals", 3]
```

+ There are 136 = 1 + 15 + 15*14/2 + 15 regression coefficients are there in this model.
+ MSE: 0.001252089
+ Yes, there are too many coefficients and only a few of them are significant, this can cause large model variance and lost of degree of freedom.

## 4b.
```{r, "forward step-wise on second order model", include = FALSE}
none_mod = lm(glyhb.star ~ 1, data=db.t)
full_mod2 = lm(glyhb.star ~ (.)^2, data=db.t)
modelfs2 = stepAIC(none_mod, scope=list(upper=full_mod2), direction="both", k=2)
```

```{r}
modelfs2$anova
```
AIC of modelfs2 = -1180.6. 
AIC of modelfs1 = -1176.62.
Best second order model under AIC criteria has smaller AIC than that of the first-order model. 
This indicates that the seconde order model is better tham first order model in model complexity and goodness of fit.

## 4c.
```{r}
plot(lm(formula = glyhb.star ~ stab.glu + ratio + age + location + 
    stab.glu:age + stab.glu:location, data = db.t), which = 1)
plot(lm(formula = glyhb.star ~ stab.glu + ratio + age + location + 
    stab.glu:age + stab.glu:location, data = db.t), which = 2)
```

Seen from the residuals vs fitted values plot, there is no non-linear pattern and basically the residuals have zero mean and equal variance. So modelfs2 is quite aduquate.

Seen from the QQ plot, the distribution of residuals is kind of heavy-tailed.

## 4d.
```{r, include = FALSE}
none_mod = lm(glyhb.star ~ 1, data=db.t)
full_mod = lm(glyhb.star ~ (.)^2, data=db.t)
stepAIC(none_mod, scope=list(upper=full_mod), direction="forward", k=2)
```

To avoid pages of tables, only the last step is shown:
```{r}
# Step:  AIC=-1180.6
# glyhb.star ~ stab.glu + ratio + age + location + stab.glu:age + 
#     stab.glu:location

#                  Df  Sum of Sq     RSS     AIC
# <none>                         0.26757 -1180.6
# + location:age    1 0.00254969 0.26502 -1180.3
# + height          1 0.00147402 0.26609 -1179.6
# + ratio:location  1 0.00088559 0.26668 -1179.2
# + time.ppn        1 0.00085521 0.26671 -1179.2
# + stab.glu:ratio  1 0.00068519 0.26688 -1179.1
# + chol            1 0.00068489 0.26688 -1179.1
# + bp.1s           1 0.00051954 0.26705 -1179.0
# + ratio:age       1 0.00051431 0.26705 -1179.0
# + hip             1 0.00042162 0.26714 -1178.9
# + bp.1d           1 0.00034397 0.26722 -1178.8
# + waist           1 0.00010658 0.26746 -1178.7
# + gender          1 0.00005509 0.26751 -1178.6
# + hdl             1 0.00001020 0.26755 -1178.6
# + weight          1 0.00000221 0.26756 -1178.6
# + frame           2 0.00179245 0.26577 -1177.8
# 
# Call:
# lm(formula = glyhb.star ~ stab.glu + ratio + age + location + 
#     stab.glu:age + stab.glu:location, data = db.t)
```


The selected model has formula:

glyhb.star ~ stab.glu + ratio + age + location + stab.glu:age + stab.glu:location

with AIC = -1180.6. It's the same as forward stepwise selection.

# 5.Model validation

## (a) Internal validation of Models fs1 and fs2

* How many regression coefficients are there in Model 3 ? 11
* What is MSE from Model 3? 0.001532671

```{r}
modelfs1
model3 = lm(glyhb.star ~ (stab.glu + ratio + age + location)^2, data = db.t)
length(model3$coefficient)
(mse_model3 = anova(model3)["Residuals", 3])
```

* Calculate SSEp, MSEp, Cp and P ressp for Models fs1 and fs2 

```{r}
#Model fs1
modelfs1 = lm(glyhb.star ~ stab.glu + ratio + age + location, data = db.t)
sse_fs1 = sum((modelfs1$residuals)^2)
mse_fs1 = anova(modelfs1)["Residuals", 3]
p_fs1 = length(modelfs1$coefficients)
n = nrow(db.t)
cp_fs1 = sse_fs1/mse_model3 - (n - 2*p_fs1)
hii = influence(modelfs1)$hat
pressp_fs1 = sum((modelfs1$residuals)^2/(1-hii)^2)
diag_fs1 = data.frame(p = p_fs1, sse = sse_fs1, mse = mse_fs1, cp = cp_fs1, pressp = pressp_fs1)

#Model fs2
modelfs2 = lm(formula = glyhb.star ~ stab.glu + ratio + age + location + stab.glu:age + stab.glu:location, data = db.t)
sse_fs2 = sum((modelfs2$residuals)^2)
mse_fs2 = anova(modelfs2)["Residuals", 3]
p_fs2 = length(modelfs2$coefficients)
n = nrow(db.t)
cp_fs2 = sse_fs2/mse_model3 - (n - 2*p_fs2)
hii = influence(modelfs2)$hat
pressp_fs2 = sum((modelfs2$residuals)^2/(1-hii)^2)
diag_fs2 = data.frame(p = p_fs2, sse = sse_fs2, mse = mse_fs2, cp = cp_fs2, pressp = pressp_fs2)

data.frame(rbind(diag_fs1, diag_fs2), row.names = c("modelfs1", "modelfs2"))
```

* Model bias in modelfs1 is quite big, with p = 5 and cp = 9.356654. This may due to unbiased estimator mse of model3?
* Model bida in modelfs2 is acceptable, with p = 7 and cp = 5.574726.
* Overfitting is not a big concern because for both models, pressp are just resonably larger than see, indicating good predictive ability.


## (b) External validation using the validation set

* Compare the coefficients:

```{r}
#Model fs1
modelfs1.v = lm(glyhb.star ~ stab.glu + ratio + age + location, data = db.v)
data.frame(rbind(modelfs1$coefficients, modelfs1.v$coefficients), row.names = c("train", "validation"))

#Model fs2
modelfs2.v = lm(glyhb.star ~ stab.glu + ratio + age + location + stab.glu:age + 
    stab.glu:location, data = db.v)
data.frame(rbind(modelfs2$coefficients, modelfs2.v$coefficients), row.names = c("train", "validation"))
```

+ For modelfs1, the coefficients from training data and validation data are consistent in sign and value. 
+ For modelfs2, other than location, all the coefficients from training data and validation data are consistent in sign and value. 
+ Yes, Models fs1 and fs2 have consistent estimates of intercept and ratio. For other coefficients, not that consistent.

* MSPE

Calculate the mean squared prediction
error (MSPE) using the validation data for each of the two models. How do these
MSP Ev compare with the respective P ressp/n and SSEp/n ()Note here n is the
sample size of the training data, i.e., 183)? Which model among the two has a
smaller MSPEv?

```{r}
#modelfs1
newdata=db.v[,-16] # the 16th is response variable glyhb.star
pred.fs1=predict.lm(modelfs1, newdata)
mspe.fs1=mean((pred.fs1-db.v[,16])^2)
mspe.fs1
pressp_fs1/nrow(db.t)
sse_fs1/nrow(db.t)
#modelfs2
newdata=db.v[,-16] # the 16th is response variable glyhb.star
pred.fs2=predict.lm(modelfs2, newdata)
mspe.fs2=mean((pred.fs2-db.v[,16])^2)
mspe.fs2
pressp_fs2/nrow(db.t)
sse_fs2/nrow(db.t)
```

+ MSEPv is supposed to be larger than sse/n. However in this case its a bit smaller than sse/n and pressp/n.
+ modelfs1 has smaller MSPEv.


## (c)

I choose modelfs1 as a final model as it has smaller MSPEv and has more degree of freedom.

```{r}
model.final = lm(glyhb.star ~ stab.glu + ratio + age + location, data = db.s)
summary(model.final)
anova(model.final)
```

# 6. Model diagnostic: Outlying and influential cases

## (a)
```{r}
plot(model.final, which = 1)
plot(model.final, which = 2)
```

+ residual vs. fitted value plot shows no nonlinear pattern, indicating that the model is sufficient.
+ residual Q-Q plot indicate that the residuals are heavy-tailed distributed.

## (b)
 Obtain the studentized deleted residuals and identify any outlying Y observations.
Use the Bonferroni outlier test procedure at ?? = 0.1.
```{r}
#studentized deleted residuals
library(MASS)
stu.res.del = studres(model.final)
head(sort(abs(stu.res.del), decreasing=TRUE))
#Bonferroni's Threshold (alpha=0.1, n=sample size, p=3)
n = nrow(db.s)
p = length(model.final$coefficients)
(qt = qt(1 - 0.1 / (2 * n), n - p - 1))
index.Y = c(195, 37, 334, 363)
```

So the first four cases all exceeds the Bonferroni's Threshold = 3.675773.
They are 195th, 37th, 334th, 363rd. 

## (c)
Obtain the leverage and identify any outlying X observations. Draw residual vs.
leverage plot.
```{r}
h = as.vector(influence(model.final)$hat)
(index.X = which(h>(2*p/n)))
length(index.X)
plot(model.final, which = 5)
```

In terms of X observations, the leverage method has detected 24 observations as outliers.

## (d)
Draw an influence index plot using Cook's distance. Are there any influential cases
according to this measure?
```{r}
plot(model.final, which = 4)
# F-test
res = model.final$residuals
mse = anova(model.final)["Residuals", 3]
cook.d = res^2*h/(p*mse*(1-h)^2)
head(sort(cook.d, decreasing = TRUE))
sort(pf(head(sort(cook.d, decreasing = TRUE)), p, n-p), decreasing = TRUE)
```

+ Even case 195, which has the largest cook's distance and p, p = 0.1113568027 < 0.2.
+ So all the cases are not influtial cases.

## (e)
Calculate the average absolute percent difference in the fitted values with and without
the most influential case identified from the previous question. What does this
measure indicate the influence of this case?

We examine the 195th case.

```{r}
model.final.influtial = lm(glyhb.star ~ stab.glu + ratio + age + location, data = db.s[-195, ])
percdif = (model.final.influtial$fitted.values - model.final$fitted.values[-195])/model.final$fitted.values[-195]
summary(percdif)
```

+ So the average absolute percent difference in the fitted values with and without the most influential case identified from the previous question is 2.855e-05, which is pretty small. 
+ This indicates that the influce of this case is not big.


